|

### 2021-02-22 supplementary_information.Rmd
Update supplementary_information.Rmd

---
title: |
  | Supplementary Information for:
  | Ecosystem size and complexity dictate riverine biodiversity
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output:
  pdf_document:
    toc: yes
    toc_depth: 2
bibliography: reference.bib
csl: nature.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

@@

$$
\begin{aligned}
u_{n} &\sim Normal(\mu_n, \sigma^2)\\
\mu_n &= \zeta_0 + \zeta_1 \sigma_{h,n} + \zeta_2 \sigma_{l,n} + \zeta_3 \sigma_{z,n} + \zeta_4 \phi_n + \zeta_5 \nu_n + \zeta_6 \alpha_{max, n} + \zeta_7 \theta_n + \zeta_8 p_{d,n} 
\end{aligned}
$$
where $\zeta_k$ ($k = 0-8$) are the intercept ($\zeta_0$) and regression coefficients ($\zeta_{1-8}$). Explanatory variables were standardized to a mean of zero and a standard deviation of one, so that regression coefficients are comparable.

@@

Longitudinal gradients of local species richness have been extensively studied in rivers, illuminating typical patterns observed in nature. The most common pattern is a downstream increase of local species richness[@mcdowallDiadromyAssemblyRestoration1996; @mcdowallFightingFlowDownstreamupstream1998; @rahelFishAssemblagesHabitat1991; @teruiThreeEcologicalFactors2016]. However, recent empirical and theoretical studies also showed 'reversed patterns,' in which local species richness decreases downstream[@bsemerHeadwatersAreCritical2013; @harveyDisturbanceReversesClassic2018]. We predicted the longitudinal gradient of local species richness to confirm that our simulation scenarios are capable of reproducing the previously observed patterns of local species richness. We considered 32 simulation scenarios comprising four landscape and eight ecological scenarios, as described in the main text (a set of parameters is described in **Table S2**). Under each simulation scenario, we generated 10 branching networks with fixed parameters of ecosystem size ($N_p = 100$) and complexity ($P_b = 0.5$). This results in a total of 320 simulation replicates. In each simulation replicate, we allowed interspecific variation in niche optimum $\mu_i$ and width $\sigma_{niche,i}$ ($\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$, respectively; subscript $i$ represents species) and ran 1400 time steps of metacommunity dynamics. We obtained temporal means of local species richness at each habitat patch for the last 1000 time steps. The first 400 time steps were discarded as initialization and burn-in periods. We removed simulation replicates in which no species established populations over the initial 400 time steps. We evaluated the relationship between local species richness and the number of upstream habitat patches, a proxy for the longitudinal position of a habitat patch.

@@

**Hokkaido, Japan.** We used data from the Hokkaido Freshwater Fish Database HFish[@fukushimaHokkaidoFreshwaterFish2011; @fukushimaModellingEffectsDams2007], monitoring data at protected watersheds[@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors[@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire Hokkaido island. Most data were collected from summer to fall. We screened data through the following procedure:

1. We listed recorded fish species and re-organized species names to make them consistent across data sources. We removed the following species at this stage: (1) identified at the family-level; (2) marine fish species (including species that occasionally use brackish/freshwater habitats).
1. We selected sampling sites based on the following criteria: (1) surveys were conducted with netting and/or electrofishing, (2) surveys were designed to collect a whole fish community, (3) sites contained reliable coordinates (sites with coordinates identical at 3 decimal degrees were treated as the same site), and (4) sites did not involve unidentified species (genus level) that are rarely observed in the data set (< 100 sites occurrence).
1. For sites with multiple visits (i.e., temporal replicates), we used the latest-year observation at each sampling site to minimize variation in sampling efforts among sites. Surveys that occurred in the same year were aggregated into a single observation.
1. We confined sites to those with the latest observation year of $\ge$ 1990. Although the data set contained observations from 1953, we added this restriction to align the observation period with the data set in the Midwest, US.
1. Four genera (*Lethenteron*, *Pungitius*, *Rhinogobius*, and *Tribolodon*) were treated as species groups (i.e., spp.) as their taxonomic resolutions varied greatly among data sources due to difficulties in identifying species.

**Midwest, US.** We assembled fish community data collected by the Iowa Department of Natural Resources[@iowadepartmentofnaturalresourcesBiologicalAssessmentIowa2004], Illinois Environmental Protection Agency and Illinois Department of Natural Resources[@illinoisenvironmentalprotectionagencyIllinoisWaterMonitoring2014], Minnesota Pollution Control Agency[@minnesotapollutioncontrolagencyDevelopmentFishbasedIndex2014], and Wisconsin Department of Natural Resources[@wisconsindepartmentofnaturalresourcesGuidelinesAssessingFish2018]. These data sets cover most of Upper Mississippi (Hydrologic Unit Code 2 [HUC 2] , region 07, as defined
by U.S. Geological Survey and U.S. Department of Agriculture Natural Resources Conservation Service[@u.s.geologicalsurveyandu.s.departmentofagriculturenaturalresourcesconservationserviceFederalStandardsProcedures2013]) and the part of Great Lakes (HUC 2, region 04), Missouri (HUC 2, region 10), and Ohio (HUC 2, region 05). Fish data were collected from summer to fall with electrofishing (backpack, barge-type, or boat-mounted) and supplemental netting at some locations. We screened data through the following procedure:

1. We used data of the Upper Mississippi (HUC 2, region 07) and Great Lakes basins (HUC 2, region 04) as most sites are included in these regions.
1. We removed records of unidentified species, hybrid species, and commercial species apparently absent in the wild (e.g., goldfish).
1. We used the latest observation at each sampling site to minimize variation in sampling efforts among sites.

@@

We evaluated sensitivity of asymptotic species richness (Chao 2 estimator) to sample size (i.e., the number of sampling sites in a watershed). We simulated presence-absence data of species with known values of true species richness $S_{true}$ and the number of sampling sites $N_{site}$. In this simulation, presence of species $i$ at site $x$, $J_{ix}$, was drawn from a Bernoulli distribution as $J_{ix} \sim Bernoulli(p_{i}\kappa_{ix})$ where $p_i$ is the detection probability for species $i$ and $\kappa_{ix}$ is the presence probability of species $i$ at site $x$. Based on the incidence frequency of simulated species $F_i = \sum_{x = 1}^{N_{site}}J_{ix}$, we estimated asymptotic species richness using the iNEXT function in the R package 'iNEXT'[@hsiehINEXTPackageRarefaction2016]. We calculated % bias of estimated asymptotic species richness $S_{est}$:

@@

The number of sampling sites influenced the estimation accuracy of asymptotic species richness. The % bias decreased sharply as the number of sampling sites increased (**Figure S20**). In particular, the estimation bias with a small sample size ($N_{site} = 5$) was substantial when the true species richness $S_{true}$ was low; some estimates showed > 150% bias (**Figure S20**). Given the simulation results, estimates of asymptotic species richness at watersheds with < 10 sampling sites may involve substantial statistical uncertainty.

@@

**Figure S19 ** Correlation structure of potential explanatory variables for riverine diversity metrics. Numeric values in each cell are the Pearson's correlation coefficients. Positive and negative correlations were colored in blue and red, respectively, and darker colors indicate stronger correlations. Environmental variables (temperature, precipitation, elevation, logit fraction of forest, logit fraction of urban, logit fraction of agriculture, and dam density) were expressed as deviations from the regional averages to remove any regional effects.

@@

**Figure S20 ** Sensitivity analysis of asymptotic species richness in relation to true species richness $S_{true}$ (panels) and the number of sampling sites $N_{site}$ (x-axis). Positive values of % bias indicate an overestimation of species richness, while negative values indicate an underestimation. Different panels show results with different true species richness. The center lines are median values, the box boundaries 25 and 75 percentiles, and the whiskers 5 and 95 percentiles. Dots represent individual simulation replicates.

>>>>

---
title: |
  | Supplementary Information for:
  | Ecosystem size and complexity dictate riverine biodiversity
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output:
  pdf_document:
    toc: yes
    toc_depth: 2
bibliography: references.bib
csl: nature.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

@@

$$
\begin{aligned}
u_{n} &\sim Normal(\mu_n, \sigma^2)\\
\mu_n &= \zeta_0 + \zeta_1 \sigma_{h,n} + \zeta_2 \sigma_{l,n} + \zeta_3 \sigma_{z,n} + \zeta_4 \phi_n + \zeta_5 \nu_n + \zeta_6 \alpha_{max, n} + \zeta_7 \theta_n + \zeta_8 p_{d,n} 
\end{aligned}
$$

where $\zeta_k$ ($k = 0-8$) are the intercept ($\zeta_0$) and regression coefficients ($\zeta_{1-8}$). Explanatory variables were standardized to a mean of zero and a standard deviation of one, so that regression coefficients are comparable.

@@

Longitudinal gradients of local species richness have been extensively studied in rivers, illuminating typical patterns observed in nature. The most common pattern is a downstream increase of local species richness[@mcdowall1996; @mcdowall1998; @rahel1991; @terui2016]. However, recent empirical and theoretical studies also showed 'reversed patterns,' in which local species richness decreases downstream[@bsemer2013; @harvey2018]. We predicted the longitudinal gradient of local species richness to confirm that our simulation scenarios are capable of reproducing the previously observed patterns of local species richness. We considered 32 simulation scenarios comprising four landscape and eight ecological scenarios, as described in the main text (a set of parameters is described in **Table S2**). Under each simulation scenario, we generated 10 branching networks with fixed parameters of ecosystem size ($N_p = 100$) and complexity ($P_b = 0.5$). This results in a total of 320 simulation replicates. In each simulation replicate, we allowed interspecific variation in niche optimum $\mu_i$ and width $\sigma_{niche,i}$ ($\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$, respectively; subscript $i$ represents species) and ran 1400 time steps of metacommunity dynamics. We obtained temporal means of local species richness at each habitat patch for the last 1000 time steps. The first 400 time steps were discarded as initialization and burn-in periods. We removed simulation replicates in which no species established populations over the initial 400 time steps. We evaluated the relationship between local species richness and the number of upstream habitat patches, a proxy for the longitudinal position of a habitat patch.

@@

**Hokkaido, Japan.** We used data from the Hokkaido Freshwater Fish Database HFish[@fukushima2011; @fukushima2007], monitoring data at protected watersheds[@comte2021; @terui2018], and primary data collected by the authors[@miyazaki2016; @terui2016], which collectively cover the entire Hokkaido island. Most data were collected from summer to fall. We screened data through the following procedure:

1.  We listed recorded fish species and re-organized species names to make them consistent across data sources. We removed the following species at this stage: (1) identified at the family-level; (2) marine fish species (including species that occasionally use brackish/freshwater habitats).
2.  We selected sampling sites based on the following criteria: (1) surveys were conducted with netting and/or electrofishing, (2) surveys were designed to collect a whole fish community, (3) sites contained reliable coordinates (sites with coordinates identical at 3 decimal degrees were treated as the same site), and (4) sites did not involve unidentified species (genus level) that are rarely observed in the data set (\< 100 sites occurrence).
3.  For sites with multiple visits (i.e., temporal replicates), we used the latest-year observation at each sampling site to minimize variation in sampling efforts among sites. Surveys that occurred in the same year were aggregated into a single observation.
4.  We confined sites to those with the latest observation year of $\ge$ 1990. Although the data set contained observations from 1953, we added this restriction to align the observation period with the data set in the Midwest, US.
5.  Four genera (*Lethenteron*, *Pungitius*, *Rhinogobius*, and *Tribolodon*) were treated as species groups (i.e., spp.) as their taxonomic resolutions varied greatly among data sources due to difficulties in identifying species.

**Midwest, US.** We assembled fish community data collected by the Iowa Department of Natural Resources[@iowadepartmentofnaturalresources2004], Illinois Environmental Protection Agency and Illinois Department of Natural Resources[@illinoisenvironmentalprotectionagency2014], Minnesota Pollution Control Agency[@minnesotapollutioncontrolagency2014], and Wisconsin Department of Natural Resources[@wisconsindepartmentofnaturalresources2018]. These data sets cover most of Upper Mississippi (Hydrologic Unit Code 2 [HUC 2] , region 07, as defined by U.S. Geological Survey and U.S. Department of Agriculture Natural Resources Conservation Service[@u.s.geologicalsurveyandu.s.departmentofagriculturenaturalresourcesconservationservice2013]) and the part of Great Lakes (HUC 2, region 04), Missouri (HUC 2, region 10), and Ohio (HUC 2, region 05). Fish data were collected from summer to fall with electrofishing (backpack, barge-type, or boat-mounted) and supplemental netting at some locations. We screened data through the following procedure:

1.  We used data of the Upper Mississippi (HUC 2, region 07) and Great Lakes basins (HUC 2, region 04) as most sites are included in these regions.
2.  We removed records of unidentified species, hybrid species, and commercial species apparently absent in the wild (e.g., goldfish).
3.  We used the latest observation at each sampling site to minimize variation in sampling efforts among sites.

@@

We evaluated sensitivity of asymptotic species richness (Chao 2 estimator) to sample size (i.e., the number of sampling sites in a watershed). We simulated presence-absence data of species with known values of true species richness $S_{true}$ and the number of sampling sites $N_{site}$. In this simulation, presence of species $i$ at site $x$, $J_{ix}$, was drawn from a Bernoulli distribution as $J_{ix} \sim Bernoulli(p_{i}\kappa_{ix})$ where $p_i$ is the detection probability for species $i$ and $\kappa_{ix}$ is the presence probability of species $i$ at site $x$. Based on the incidence frequency of simulated species $F_i = \sum_{x = 1}^{N_{site}}J_{ix}$, we estimated asymptotic species richness using the iNEXT function in the R package 'iNEXT'[@hsieh2016]. We calculated % bias of estimated asymptotic species richness $S_{est}$:

@@

The number of sampling sites influenced the estimation accuracy of asymptotic species richness. The % bias decreased sharply as the number of sampling sites increased (**Figure S20**). In particular, the estimation bias with a small sample size ($N_{site} = 5$) was substantial when the true species richness $S_{true}$ was low; some estimates showed \> 150% bias (**Figure S20**). Given the simulation results, estimates of asymptotic species richness at watersheds with \< 10 sampling sites may involve substantial statistical uncertainty.

@@

**Figure S19** Correlation structure of potential explanatory variables for riverine diversity metrics. Numeric values in each cell are the Pearson's correlation coefficients. Positive and negative correlations were colored in blue and red, respectively, and darker colors indicate stronger correlations. Environmental variables (temperature, precipitation, elevation, logit fraction of forest, logit fraction of urban, logit fraction of agriculture, and dam density) were expressed as deviations from the regional averages to remove any regional effects.

@@

**Figure S20** Sensitivity analysis of asymptotic species richness in relation to true species richness $S_{true}$ (panels) and the number of sampling sites $N_{site}$ (x-axis). Positive values of % bias indicate an overestimation of species richness, while negative values indicate an underestimation. Different panels show results with different true species richness. The center lines are median values, the box boundaries 25 and 75 percentiles, and the whiskers 5 and 95 percentiles. Dots represent individual simulation replicates.
